Grade Retention and Mixed Race

### 2022-09-07 products/paper/main.qmd
Edit biracial results section

I focus first on biracial groups with one Black parent because of a stronger historical expectation that these students will be identified as Black by the "one-drop rule" and as a result will have similar outcomes to Black students. In three cases, this hypothesis can be soundly rejected. For Black/White, Black/Latino, and Black/Asian students, The probability of being clearly below expected grade is substantially lower than that for monoracial black students and the difference is statistically distinguishable from zero. Only in the case of Black/Indigenous students, do I observe outcomes that are indistinguishable from those of Black students. However, in this case,monoracial Black students have lower risk of grade retention than monoracial Indigenous students. Thus, In all four cases, part-Black biracial students have a risk of grade retention that is closer to the lower risk monoracial group.

The results also suggest that each of these biracial groups has a risk of grade retention lower than the halfway point, although this hypothesis cannot be rejected completely for the Black/Asian and Black/Indigenous groups, due to a higher degree of uncertainty for these smaller groups.

Controlling for resource variables has some effect on the placement of Black/White and Black/Latino biracial respondents. In the case of Black/White biracial respondents, Black/White students initially have a risk of grade retention slightly higher than and statistically distinguishable from White students. After controlling for cultural and family resources, however, Black/White respondents have a risk of grade retention that is slightly lower and statistically distinguishable from White students. This change reflects the relatively low level of parental education and income among Black/White students, which is much closer to Black students than White students, as shown on @tbl-descriptive.

An opposite pattern is observed for Black/Latino respondents. Black/Latino respondents initially have a risk of grade retention substantially lower than both Black and Latino students, who have similar risks. However, after controlling for cultural and family resources, Black/Latino students have a risk of grade retention that is indistinguishable from the grade retention risk of Latino students and substantially lower than Black students. The change largely reflects the fact that Black/Latino students have greater cultural resources than Latino students and once this advantage is accounted for, they have a similar risk.

Altogether, the results here strongly suggest that the risk of grade retention faced by these biracial students with one black parent is generally similar to that of their constituent monoracial group with the lower risk of grade retention. This finding is inconsistent both with the expectation of the one-drop rule and the "in-between" hypothesis.

@fig-other-models shows the results for the remaining biracial groups. In general, the results here are consistent with those for part-Black biracial respondents, with some important caveats discussed below. The point estimates indicate that the risk of each biracial group is either quite close to the monoracial group with the lower risk or in-between this group's risk and the halfway point.

@@

For all groups, the full models generally indicate that each biracial group has a risk of grade retention similar to the risk of the lower monoracial constituent group or has risk higher than this group but less than the halfway point. There are two exceptions to this general pattern. First, for the Indigenous/Latino group, I cannot rule out the possibility that this group has risk roughly halfway between the two constituent groups, although the point estimate favors a result closer to the lower Latino group. Second, Latino/Asian students are the only case where the risk for the biracial group is higher than either monoracial group. However, the confidence bands suggest that the risk for the Latino/Asian group cannot be differentiated from the rates for Latinos, the higher of the two monoracial constituent groups. This result marks the one substantial deviation from the general pattern. It only emerges in the models that control for cultural and family resources because Latino/Asian students come from higher income and more educated households than Latino students, on average.

Aside from the effect on Latino/Asian students, controlling for cultural and family resources has moderate effects on the placement of the remaining biracial groups. The only case with a substantial shift is that for White/Asian students who initially appear to have lower risk than either White or Asian students. After controlling for these resources, the risks of White/Asian students are indistinguishable from the lower risk of Asian students. This is due to the fact that White/Asian students come from the most advantaged households of any group in the dataset.

Any results for part-Latino students are complicated by the substantial change in the monoracial Latino risk across models. For example, while Latinos have substantially higher risk of grade retention than Whites in the baseline model, there is only a very small difference favoring Latinos in the full model. Nonetheless, even in this case, we see that White/Latinos have risk closer to those of the lower Latino group, but distinguishable from both groups.

>>>>

I focus first on biracial groups with one Black parent because of a stronger historical expectation that these students will be identified as Black by the "one-drop rule" and as a result will have similar outcomes to Black students. @fig-black-models provides little evidence for such a pattern, and soundly rejects it in the case of Black/White and Black/Latino students. Black/White and Black/Latino students have a risk of BBEG substantially lower than Black students and closer, and perhaps even lower, than their non-Black constituent group. Solid conclusions are difficult for the remaining two groups because of smaller sample size and correspondingly wider confidence intervals on estimates. Nonetheless the point estimates in each of these cases suggest probabilities of BBEG roughly midway between the two constituent groups.

Controlling for resource variables has some effect on the placement of part-Black biracial respondents. In the case of Black/White biracial respondents, Black/White students initially have a risk of BBEG slightly higher than and statistically distinguishable from White students. After controlling for cultural and family resources, however, Black/White respondents have a risk of BBEG that is slightly lower and statistically distinguishable from White students. This change reflects the relatively low level of parental education and income among Black/White students, which is much closer to Black students than White students, as shown on @tbl-descriptive.

An opposite pattern is observed for Black/Latino respondents. Black/Latino respondents initially have a risk of grade retention substantially lower than both Black and Latino students, who have similar risks. However, after controlling for cultural and family resources, Black/Latino students have a risk of BBEG that is indistinguishable from the risk of Latino students and substantially lower than Black students. The change largely reflects the fact that Black/Latino students have greater resources than Latino (but not Black) students and once this advantage is held constant, they have a similar risk.

The role of parental resources is harder to determine for Black/Asian and Black/Indigenous students due to the greater uncertainty in these estimates. Controlling for resources moves the point estimate for Black/Asian students from a point estimate close to Asian students to more of a halfway position between the two constituent groups. Because Black/Indigenous students had resources halfway between their constituent groups, controlling for these resources hardly affects the relative placement of Black/Indigenous students at all.

@fig-other-models shows the results for the remaining biracial groups. In general, the results here show that biracial students are more likely to have a risk of BBEG more like the monoracial constituent group with the lower risk than the higher risk, with some important caveats discussed below.

@@

For all groups, the full models generally indicate that each biracial group has a risk of BBEG similar to the risk of the lower monoracial constituent group or has risk higher than this group but less than the halfway point. There are two exceptions to this general pattern. First, the point estimate of the probability of BBEG for the Indigenous/Latino group is roughly halfway between the two constituent groups and statistically distinguishable from both. Second, Latino/Asian students are the only case where the point estimate of the risk for the biracial group is higher than either monoracial group. However, the confidence bands suggest that the risk for this group cannot be statistically differentiated from either of the two monoracial constituent groups. This result only emerges in the model that controls for resources, Latino/Asian students come from more advantaged households than both Latino and Asian students, on average.

Aside from the effect on Latino/Asian students, controlling for resources has moderate effects on the placement of the remaining biracial groups. White/Asian students are the only case with a substantial shift, owing to their highly advantaged households. After controlling for resources, the risks of BBEG for White/Asian students are indistinguishable from the risk of Asian students, but still substantially and statistically distinguishable from the higher risk to White students.

Results for part-Latino students are complicated by the substantial change in the monoracial Latino risk across models. For example, while Latinos have substantially higher risk of BBEG than Whites in the baseline model, there is only a very small difference favoring Latinos in the full model. Nonetheless, even in this case, we see that White/Latinos have risk closer to those of the lower Latino group, but distinguishable from both groups.
